---
title: "Objective 3"
output: html_document
date: "2025-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  CHAPTER 1: Introduction
## 1.0 Overview

The spatial pattern of fire incidents is a critical area of study, particularly in urban environments where population density and infrastructure can significantly influence fire dynamics. This research project focuses on the spatial distribution of fire incidents in Washington, D.C., Maryland, and Virginia, examining how geographic and environmental factors contribute to the occurrence and frequency of fires.

## 1.1 Objectives

Analyze the spatial distribution of fire incidents across Washington, D.C., Maryland, and Virginia, identifying geographic hotspots and patterns in urban versus rural areas.
Investigate the role of environmental factors, such as weather conditions and vegetation, in influencing the incidence of fires in the studied regions during 2018.
Categorize types of fire incidents (e.g., residential, commercial, vehicle) and assess their causes, comparing urban and rural settings to identify specific risk factors.

## 1.2 Research Questions

What geographic patterns can be identified in the distribution of fire incidents across Washington, D.C., Maryland, and Virginia in 2018?

How do environmental factors influence the incidence of fires in the studied regions during 2018?

What types of fire incidents were most prevalent, and what are the specific causes associated with these incidents in both urban and rural settings?

## 1.3 Significance of the Study
This project study enhances understanding of fire incident dynamics in Washington, D.C., Maryland, and Virginia. Findings can inform policymakers and emergency services about high-risk areas and the underlying causes of fires, facilitating targeted prevention strategies and improving public safety initiatives.

#  CHAPTER 2: Methodology
## 2.0 Dataset 
This project utilized the US Wildfire Dataset from Kaggle, comprising approximately 1.9 million observations of wildfire incidents across the United States. For this analysis, approximately 17,000 fire incidents from D.C., Maryland, and Virginia in 2018 were filtered from the main dataset.

spatial data for project study area (DC,MD AND VA) were downloaded from Diva GIS (https://diva-gis.org/data.html)

## 2.1 Software
The following software was used for this project:

R and R Markdown
GitHub (GitHub Repository)
Microsoft Excel

## 2.2 Project Study Area
# Objective 3: Categorize Types of Fire Incidents and Assess Causes



```{r init, include=F}
library(ezids)
library(ggplot2)
library(treemapify)
```


# Read dataset
```{r}
fire <- read.csv("Fire incidences DC_VA_MD_cleands.csv")
```


# Quick overview
```{r}
str(fire)
head(fire)
```


# summary 
```{r}
xkablesummary(fire, title = "Fire Statistics Summary")
```

#1. CATEGORIZATION: Fire Incident Types and Causes

```{r}
table(fire$NWCG_CAUSE_CLASSIFICATION)
```


# Bar Chart
```{r}
ggplot(fire, aes(x = NWCG_CAUSE_CLASSIFICATION)) +
  geom_bar(col = "black", fill = "white", alpha = 0.7) +
  labs(title = "Fire Cause Classification",
       x = "Cause Classification",
       y = "Number of Incidents")
```

# Frequency count of causes
```{r}
general_cause_counts <- table(fire$NWCG_GENERAL_CAUSE)
general_cause_counts <- sort(general_cause_counts, decreasing = TRUE)
general_cause_counts
```

```{r}
fire_cause_summary <- as.data.frame(table(fire$NWCG_GENERAL_CAUSE))
colnames(fire_cause_summary) <- c("Cause", "Count")
ggplot(fire_cause_summary, aes(area = Count, fill = Cause, label = paste(Cause, "\n", Count))) +
  geom_treemap() +
  geom_treemap_text(colour = "white", place = "centre", grow = TRUE) +
  labs(title = "Fire Incident Causes",
       subtitle = "Proportion of Each General Cause across All Regions") +
  theme(legend.position = "none")
```
# Segregation based on states
```{r}
table(fire$STATE)
```

# Bar chart on fire incidences in States.
```{r}
state_counts <- table(fire$STATE)
barplot(state_counts,
        main = "Fire Incidents by State",
        xlab = "State",
        ylab = "Number of Incidents",
        col = "red")
``` 

# Classification of urban and rural counties in DC,MD and VA.
```{r}
urban_counties <- c("Washington, D.C.",
                    "Montgomery County", "Prince George's County", "Baltimore City",
                    "Anne Arundel County", "Howard County",
                    "Fairfax County", "Arlington County", "Virginia Beach City",
                    "Chesterfield County", "Henrico County", "Loudoun County")

rural_counties <- c("Garrett County", "Allegany County", "Washington County",
                    "Caroline County", "Somerset County", "Dorchester County",
                    "Bath County", "Highland County", "Scott County",
                    "Lee County", "Tazewell County", "Wise County")

# Create new Area Type variable
fire$AREA_TYPE <- ifelse(fire$FIPS_NAME %in% urban_counties, "Urban", "Rural")

# View distribution
table(fire$AREA_TYPE)
```
```{r}
ggplot(fire, aes(x = AREA_TYPE, y = FIRE_SIZE, fill = AREA_TYPE)) +
  geom_boxplot(alpha = 0.7, color = "black", outlier.colour = "red") +
  labs(title = "Fire Size Distribution by Area Type",
       x = "Area Type",
       y = "Fire Size") +
  scale_fill_manual(values = c("blue", "orange", "black")) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 0, hjust = 0.5))
```
# Frequency table of causes by area type
```{r}
cause_table <- table(fire$AREA_TYPE, fire$NWCG_GENERAL_CAUSE)
cause_table
```
# Distibution of fire cause by area type
```{r}
ggplot(data = fire, aes(x = NWCG_GENERAL_CAUSE, fill = AREA_TYPE)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Fire Causes by Area Type",
       x = "General Cause", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Identify risk factors associated with different areas.
```{r}
risk_factor <- chisq.test(table(fire$AREA_TYPE, fire$NWCG_GENERAL_CAUSE))
risk_factor
```
Fire causes differs significantly across area types (Urban, Rural, Other).
# Normality Check for Fire Size
```{r}
qqnorm(fire$FIRE_SIZE, main = "Fire Size") ## Assess whether the distribution of fire sizes follows a normal distribution.
qqline(fire$FIRE_SIZE)
```
