---
title: "Object 2"
author: "Jeongwon Yoo"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: 3
---
  
```{r setup, include=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install required packages if they are not already installed
required_packages <- c("sf","ggplot2","viridis","dplyr","lubridate",
                       "corrplot","tigris","readr","tidyr",
                       "spatialreg","spdep")
new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]
if (length(new_packages)) {
  install.packages(new_packages)
}

# Load libraries
library(sf)
library(ggplot2)
library(viridis)
library(dplyr)
library(lubridate)
library(tigris)
library(readr)
library(tidyr)
library(corrplot)
library(spatialreg)
```


```{r include=FALSE} 
# Set global chunk options
knitr::opts_chunk$set(echo = FALSE, results = "markup", warning = FALSE, message = FALSE)

# Set options for number display format
options(scientific = TRUE, digits = 3)
```

# Object 2

Investigate Environmental Factors Influencing Fire Incidences-Jeongwon Yoo

## 1. Data Integration

```{r}
# Read CSV
clean_df = '/Users/jeongwonyoo/Documents/GitHub/introds-project1/Fire incidences DC_VA_MD_cleaned.csv' 
raw_df = '/Users/jeongwonyoo/Documents/GitHub/introds-project1/Wildfire_Dataset.csv'

library('readr')
clean_df = read_csv(clean_df)
raw_df = read_csv(raw_df)

# Make row_df columns names same as clean_df
names(raw_df) = toupper(names(raw_df))

# Add State name on raw data
options(tigris_use_cache = TRUE)

states <- tigris::states(cb = TRUE, year = 2023) %>%
    st_transform(4326) %>%
    select(STUSPS, NAME, geometry)

sf_raw <- st_as_sf(raw_df %>% filter(!is.na(LATITUDE), !is.na(LONGITUDE)),coords = c("LONGITUDE","LATITUDE"), crs = 4326, remove=FALSE)

raw_with_state_poly <- st_join(sf_raw, states, join = st_within) %>%
    st_drop_geometry() %>%
    rename(STATE = STUSPS)

# sort target data(state(VA,DC,MD),time(2018))
target_states <- c("DC","MD","VA")
target_year   <- 2018
cols_keep     <- c("STATE","FIRE_YEAR","Month","LATITUDE","LONGITUDE",
                   "DATETIME","PR","RMAX","RMIN","SPH","SRAD","TMMN",
                  "TMMX","VS","BI","FM100","FM1000","ERC","ETR","PET","VPD")

# Make sub dataset of clean_df
clean_sub <- clean_df %>%
    filter(STATE %in% target_states,
           FIRE_YEAR == target_year) %>% 
    select(any_of(cols_keep))

# Make sub dataset of raw_df
raw_sub <- raw_with_state_poly %>%
    filter(STATE %in% target_states,
           lubridate::year(lubridate::as_datetime(DATETIME)) == target_year) %>%
    drop_na(LATITUDE, LONGITUDE) %>%
    select(any_of(cols_keep))

# organized data check
str('raw_sub')
```

Call new dataset for the analysis
```{r}
dp = '/Users/jeongwonyoo/Documents/GitHub/introds-project1/clean_object2_2018_DCMDVA.csv'
object2 = read_csv(dp)
```

## 2. Correlation Analysis
Correlation Analysis

```{r}
env_vars <- object2 %>% select(PR, RMAX, RMIN, SPH, SRAD, TMMN, TMMX, VS, FM100, FM1000, ERC, ETR, PET, VPD)
cor_mat <- cor(env_vars, use = "complete.obs")
corrplot(cor_mat, method = "color", tl.cex = 0.8)
```

Figure 1. Correlation heatmap.
The heatmap shows strong negative correlations between ERC and fuel moisture variables (FM100, FM1000), indicating that wetter fuels are associated with lower fire potential. ERC is moderately positively correlated with atmospheric dryness and heat (VPD, TMMX). Humidity indicators (RMAX, RMIN, SPH) are strongly negatively correlated with VPD, which is expected since higher humidity coincides with lower vapor pressure deficit. FM100 and FM1000 are highly positively correlated, so they should not be used together in the same model due to multicollinearity risk.

### 2.1. Raster Maps

Figure: Maximum Temperature Distribution (°C)
```{r}
library(dplyr); library(ggplot2)

plot_raster <- function(df, var, bins = 60, to_celsius = FALSE) {
    z <- if (to_celsius) df[[var]] - 273.15 else df[[var]]
    ggplot(df, aes(LONGITUDE, LATITUDE, z = z)) +
        stat_summary_2d(fun = mean, bins = bins, na.rm = TRUE) +
        coord_fixed(expand = FALSE) +
        scale_fill_viridis_c(name = var) +
        labs(title = paste(var, "Distribution")) +
        theme_minimal(base_size = 12)
}
```

```{r}
plot_raster(object2, "TMMX", bins = 70, to_celsius = TRUE)
```

The binned raster map shows a west-to-east warming gradient, with higher maximum temperatures concentrated in the eastern and southeastern bins and cooler conditions in the western highlands, consistent with regional topography and supporting the observed positive association between heat and fire potential.

```{r}
plot_raster(object2, "VPD",  bins = 70)
```

```{r}
plot_raster(object2, "FM100", bins = 70)
```

```{r}
plot_raster(object2, "PR",   bins = 70)
``` 

```{r}
plot_raster(object2, "VS",   bins = 70)
```

```{r}
plot_raster(object2, "ERC",  bins = 70)
```

### 2.2. Scatter Plots

Scatter Plots

```{r}
ggplot(object2, aes(x = TMMX, y = BI)) +geom_point(alpha = 0.4)+     geom_smooth(method = "lm", se = FALSE, color = "red")+labs(title = "Fire Index vs. Temperature")
```

Figure 2. BI vs. TMMX.
The fitted line shows a weak relationship and the distribution is dominated by many zero or near-zero BI values. This zero inflation and visible heteroscedasticity limit the usefulness of simple bivariate OLS. A multivariable model that controls for dryness (VPD), fuel moisture (FM100), wind (VS), and precipitation (PR) is more appropriate. If the goal is to model BI directly, consider restricting to BI > 0 or using a model that handles zero inflation or nonlinearity.

### 2.3 Regression Diagnostics Plots

Regression Diagnostics Plots

```{r}
model <- lm(BI ~ TMMX + PR + VS + VPD, data = object2)
par(mfrow = c(2, 2))
plot(model)  # residuals, Q-Q, leverage plots
```

Figure 3. OLS diagnostic plots.
Residuals vs. fitted values display a fan shape and curvature, which indicates heteroscedasticity and potential nonlinearity. The Q–Q plot shows tail deviations, suggesting heavy-tailed residuals and outliers. The scale-location plot confirms nonconstant variance. The residuals-versus-leverage plot highlights a few influential observations. These patterns show that standard OLS assumptions are not well met and motivate spatial regression or robust alternatives.

### 2.4 Spatial Regression

```{r}
# Build spatial dataset and weights for modeling
# Collapse to one record per location to avoid duplicate coordinates by date
o2 <- object2 |>
  dplyr::group_by(LONGITUDE, LATITUDE) |>
  dplyr::summarise(dplyr::across(c(PR, RMAX, RMIN, SPH, SRAD, TMMN, TMMX, VS,
                                   FM100, FM1000, ERC, ETR, PET, VPD),
                                 ~mean(.x, na.rm = TRUE)),
                   .groups = "drop")

#| label: spatial-weights
vars <- c("ERC","TMMX","VPD","FM100","PR","VS","RMAX","SRAD")
dat  <- na.omit(o2[, vars, drop = FALSE])

coords_all <- as.matrix(o2[, c("LONGITUDE","LATITUDE")])
coords_k   <- coords_all[as.numeric(rownames(dat)), , drop = FALSE]

kn1  <- spdep::knearneigh(coords_k, k = 1, longlat = TRUE)
dmax <- max(kn1$nn.dist)
nb_k <- spdep::dnearneigh(coords_k, d1 = 0, d2 = dmax * 1.05, longlat = TRUE)

while (any(spdep::card(nb_k) == 0)) {
  dmax <- dmax * 1.2
  nb_k <- spdep::dnearneigh(coords_k, d1 = 0, d2 = dmax, longlat = TRUE)
}
if (spdep::n.comp.nb(nb_k)$nc > 1) {
  k <- 5
  repeat {
    nb_k <- spdep::knn2nb(spdep::knearneigh(coords_k, k = k, longlat = TRUE))
    if (spdep::n.comp.nb(nb_k)$nc == 1 || k >= 20) break
    k <- k + 1
  }
}

lw_k <- spdep::nb2listw(nb_k, style = "W", zero.policy = FALSE)
```

```{r}
# Fit models
ols <- lm(ERC ~ TMMX + VPD + FM100 + PR + VS + RMAX + SRAD, data = dat)
sar <- lagsarlm(ERC ~ TMMX + VPD + FM100 + PR + VS + RMAX + SRAD,
                data = dat, listw = lw_k, zero.policy = TRUE, na.action = na.exclude)
sem <- errorsarlm(ERC ~ TMMX + VPD + FM100 + PR + VS + RMAX + SRAD,
                  data = dat, listw = lw_k, zero.policy = TRUE, na.action = na.exclude)

# Summaries
summary(ols); summary(sar); summary(sem); AIC(sar, sem)

# Spatial autocorrelation of residuals
spdep::lm.morantest(ols, lw_k, zero.policy = TRUE)         # OLS
moran.test(residuals(sar), lw_k, zero.policy = TRUE)        # SAR
moran.test(residuals(sem), lw_k, zero.policy = TRUE)        # SEM
```

Residual spatial autocorrelation after SEM.
Moran’s I on the SEM residuals is not significant (I ≈ −0.059, z ≈ −0.76, p ≈ 0.78). This indicates that the spatial error model adequately removed spatial clustering in the residuals. In practical terms, the spatial structure is captured by the model and remaining errors are close to spatially random.

What this implies for drivers of fire activity.
Taken together, the figures are consistent with the following: hot and dry conditions (higher TMMX and VPD) are associated with increased fire potential, while higher fuel moisture and wetter conditions (FM100, FM1000, PR, higher humidity) are associated with reduced fire potential. Because FM100 and FM1000 are strongly collinear, use only one of them in the final model. Report SAR vs. SEM selection using AIC, then present the signs and significance of the key coefficients, followed by the non-significant Moran’s I on residuals to demonstrate that model assumptions are satisfied.

# last edit by simba
# added other plots and finialising it(Jeongwon)